Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-05 numerical-ppm.tex
Edits to the overview section. Trying to unify notations, so several other sections are also affected.

To overcome this, we also solve the internal energy equation:
\begin{equation}
 \frac{\partial e}{\partial t} 
           + \frac{1}{a} \vecv \cdot \nabla e
%         = - \frac{3(\gamma -1)\dot{a}}{a} e
         = - 3 \frac{\dot{a}}{a} \frac{p}{\rho}
           - \frac{p}{a\rho} \nabla \cdot \vecv
\end{equation}
in comoving coordinates.  The structure is similar to the total energy
equation; the second term on the left hand side represents transport, while
the first term on the right is due to expansion of the coordinate
system.  
It is differenced (again, in Eulerian form without the expansion term) as,
%
\begin{equation}
\rho_j^{n+1} e_j^{n+1}  = 
       \rho_j^n e_j^n   +  \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{e}  - \avm{\rho} \avm{v} \avm{e}} {\Delta x_j} \right)
            - \Delta t \ p_j^{n} \left( \frac{ \avp{v} - \avm{v}} {\Delta x_j} 
                      \right)
                  \label{eq:gasenergy_diff}
\end{equation}
%
Note that because of the structure of this equation, it is not in
flux-conservative form.  In particular, the pressure is evaluated at
the cell center.  Unfortunately, time-centering of this pressure has
proved difficult to do without generating large errors in the internal
energy and so we conservatively leave the pressure at the old time in
this difference equation.  This leads to spreading of shocks; however,
we note that this equation is only used in hypersonic flows.

@@

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be found by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in Enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works with both
MUSCL and MHD-CT methods.

@@

For the MHD equations, there are seven waves, instead of three.  This makes the
exact solution to the Riemann quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
Dedner and CT solvers in Enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.

@@

%\begin{equation}
%\frac{\partial e}{\partial t} + \frac{1}{a} \vecv_b \cdot \nabla e = 
%- \frac{3(\gamma-1)\dot{a}}{a}e - \frac{p}{a \rho} \nabla \cdot \vecv_b
%\label{enzointenergy}
%\end{equation}

>>>>

To overcome this, we also solve the internal energy equation:
\begin{equation}
 \frac{\partial e}{\partial t} 
           + \frac{1}{a} \vecv \cdot \grad e
%         = - \frac{3(\gamma -1)\dot{a}}{a} e
         = - 3 \frac{\dot{a}}{a} \frac{p}{\rho}
           - \frac{p}{a\rho} \div \vecv
\end{equation}
in comoving coordinates.  The structure is similar to the total energy
equation; the second term on the left hand side represents transport, while
the first term on the right is due to expansion of the coordinate
system.  
It is differenced (again, in Eulerian form without the expansion term) as,
%
\begin{equation}
\rho_j^{n+1} e_j^{n+1}  = 
       \rho_j^n e_j^n   +  \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{e}  - \avm{\rho} \avm{v} \avm{e}} {\Delta x_j} \right)
            - \Delta t \ p_j^{n} \left( \frac{ \avp{v} - \avm{v}} {\Delta x_j} 
                      \right)
                  \label{eq:gasenergy_diff}
\end{equation}
%
Note that because of the structure of this equation, it is not in
flux-conservative form.  In particular, the pressure is evaluated at
the cell center.  Unfortunately, time-centering of this pressure has
proved difficult to do without generating large errors in the internal
energy and so we conservatively leave the pressure at the old time in
this difference equation.  This leads to spreading of shocks; however,
we note that this equation is only used in hypersonic flows.

@@

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be found by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in \enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works with both
MUSCL and MHD-CT methods.

@@

For the MHD equations, there are seven waves, instead of three.  This makes the
exact solution to the Riemann quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
Dedner and CT solvers in \enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.

@@

%\begin{equation}
%\frac{\partial e}{\partial t} + \frac{1}{a} \vecv_b \cdot \grad e = 
%- \frac{3(\gamma-1)\dot{a}}{a}e - \frac{p}{a \rho} \div \vecv_b
%\label{enzointenergy}
%\end{equation}

### 2013-06-05 numerical-radiation-fld.tex
Edits to the overview section. Trying to unify notations, so several other sections are also affected.

Enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \nabla\cdot\left(E_r \vec{v}_b\right) &=& 
%    \nabla\cdot\left(D\nabla E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \nabla\cdot\left({\mathrm n}_i \vec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(Equation~\ref{eq:species_evolution}) is solved
independently of Enzo's hydrodynamics, gravity and dark-matter solvers
(Sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of Equations (\ref{eq:fld_radiation}) and
(\ref{eq:species_evolution}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given time step to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

>>>>

\enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \div \left(E_r \vec{v}_b\right) &=& 
%    \div\left(D\grad E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \vec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(Equation~\ref{eq:species_evolution}) is solved
independently of \enzo's hydrodynamics, gravity and dark-matter solvers
(Sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of Equations (\ref{eq:fld_radiation}) and
(\ref{eq:species_evolution}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given time step to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

### 2013-06-05 numerical-timestep.tex
Edits to the overview section. Trying to unify notations, so several other sections are also affected.

\begin{equation}
\Delta t_{\rm cond} = \min \left(  \frac{ \kappa_{\rm cond}}{f_{\rm Sp}} \frac{\Delta x^2
    n_b}{\kappa(T)} \right)_L,
\label{eqn:dtcond}
\end{equation}

@@

Equation~(\ref{eqn:dtcond}) is the stability condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density. 
%defined as $n_b = \rho_b / \mu
%m_p$, where $\mu$ is the mean molecular weight, 
$\kappa(T)$ is the Spitzer thermal conductivity, and f$_{Sp}$ is a user-defined
conduction suppression factor whose value must be f$_{Sp} \leq 1$.  
%Both are defined in Section~\ref{sec.num.conductions}.  
$\kappa_{\rm cond}$ is a prefactor whose
value must be $0 < \kappa_{\rm cond}  < 0.5$, and is exactly 0.5 for the
implemention in \enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$),
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.

>>>>

\begin{equation}
\Delta t_{\rm cond} = \min \left(  \frac{ \kappa_{\rm cond}}{f_{\rm sp}} \frac{\Delta x^2
    n_b}{\kappa(T)} \right)_L,
\label{eqn:dtcond}
\end{equation}

@@

Equation~(\ref{eqn:dtcond}) is the stability condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density. 
%defined as $n_b = \rho_b / \mu
%m_p$, where $\mu$ is the mean molecular weight, 
$\kappa(T)$ is the Spitzer thermal conductivity, and f$_{\rm sp}$ is a user-defined
conduction suppression factor whose value must be f$_{\rm sp} \leq 1$.  
%Both are defined in Section~\ref{sec.num.conductions}.  
$\kappa_{\rm cond}$ is a prefactor whose
value must be $0 < \kappa_{\rm cond}  < 0.5$, and is exactly 0.5 for the
implemention in \enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$),
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.
